A formal treatment of bidirectional typing\texorpdfstring{\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}}{}

### 2023-12-25 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Add rule names for annotation ordering
[ tex ] Add the revised remark about annotatability

\begin{figure}
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}
              {\isTerm{x} \sqsupseteq \isTerm{x}}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq (\isTerm{u} \annotate A)}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
              {\tmOpts \sqsupseteq \tmOpus}
  \end{mathpar}
  \caption{Annotation ordering between raw terms}
  \label{fig:annotation-ordering}
\end{figure}
On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), our completeness is much simpler to use than annotatability, which requires the bidirectional type synthesiser to produce more complex evidence when the synthesis fails.
Annotatability also does not help the user to deal with missing annotations like (generalised) mode decoration does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
%Therefore we believe that the notion of annotatability is too complex\LT{not clear what it means} for what it achieves, and our theory offers simpler and more useful alternatives.

>>>>

\begin{figure}
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}{\isTerm{x} \sqsupseteq \isTerm{x}}\;\Rule{Var}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
      {(\isTerm{t} \annotate A) \sqsupseteq (\isTerm{u} \annotate A)}\;\Rule{Anno}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
      {\tmOpts \sqsupseteq \tmOpus}\;\Rule{Op}
  \end{mathpar}
  \caption{Annotation ordering between raw terms}
  \label{fig:annotation-ordering}
\end{figure}

On the other hand, when using a bidirectional type synthesiser to implement a type synthesiser, for example in \cref{thm:implementation}, if the bidirectional type synthesiser concludes that there does not exist a bidirectional typing derivation, we use the contrapositive form of completeness to establish that such an ordinary typing derivation does not exist either.
Now, annotatability is a kind of completeness because (roughly speaking) it turns an ordinary typing derivation bidirectional.
Therefore it is conceivable that we could use annotatability in place of completeness in the proof of \cref{thm:implementation}.
However, in the contrapositive form of annotatability, the antecedent is `there does not exist $t'$ that is more annotated than $t$ and has a bidirectional typing derivation', which is more complex than the bidirectional type synthesiser would have to produce.
Annotatability also does not help the user to deal with missing annotations: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
Therefore we believe that our theory offers simpler and more useful alternatives than the notion of annotatability.
